---
title: "EDA_GroupG"
output: html_document
date: "2025-10-13"
Authors: "Oksana Efron, Sabrina Hassan, Keri Gagnow"
---

```{r setup, include=FALSE}
library(readr)
library(psych)
library(corrplot)
library(ggplot2)
```
Importing our dataset into R Markdown:
```{r}
student_depression_dataset <- read_csv("student_depression_dataset_CSV.csv")
head(student_depression_dataset)
```
We found a data set that analyzes mental health trends and predictors among students, which we ended up naming to StudentDep. We reorganized the data so it has 70 cases and 17 variables. Our goal was to make sure that the variables was layed out properly and in a better fit so we could analyze.

```{r}
StudentDep <- student_depression_dataset[1:70, c("Gender", "Age", "City", "Profession", "Academic Pressure", "Work Pressure", "CGPA", "Study Satisfaction", "Job Satisfaction", "Sleep Duration", "Dietary Habits", "Degree", "Have you ever had suicidal thoughts ?", "Work/Study Hours", "Financial Stress", "Family History of Mental Illness", "Depression")]
dim(StudentDep)
```

```{r}
StudentDep
```
What we did here is converted the categorical variables into numerical form, such as 1 for males and 0 for females. As well as for Dietary Habits and Sleep Duration, they were changed into numeric factors.

```{r}
# Making sure to change Gender to binary so its Males = 1, and Female = 0
StudentDep$Gender <- ifelse(StudentDep$Gender == "Male", 1, 0)

# Also doing this for Family History of Mental Illness converting, binary (Yes = 1, No = 0)
StudentDep$'Family History of Mental Illness' <- ifelse(StudentDep$'Family History of Mental Illness' == "Yes", 1, 0)

# Converting ordinal variables for Dietary Habits and Sleep Duration
StudentDep$ 'Dietary Habits' <- as.numeric(factor(StudentDep$ 'Dietary Habits',
                                                  levels = c("Unhealthy", "Average", "Healthy"),
                                                  ordered = TRUE))
StudentDep$ 'Sleep Duration' <- as.numeric(factor(StudentDep$ 'Sleep Duration',
                                                  levels = c("<5 hours", "5-6 hours", "6-8 hours", ">8 hours"),
                                                  ordered = TRUE))

# Converting Depression, for binary levels. Yes = 1 and No = 0
StudentDep$Depression <- ifelse(StudentDep$Depression == "Yes", 1, 0)

str(StudentDep)
```
```{r}
summary(StudentDep)
```
```{r}
dim(StudentDep)
```
For here we made sure to have variables that are predictors, and variables that are response variables.

```{r}
# Predictor Variables
X <- StudentDep[, c("Gender", "Age", "Academic Pressure", "Work Pressure",
                    "CGPA", "Dietary Habits", "Financial Stress",
                    "Family History of Mental Illness")]
# Response Variables
Y <- StudentDep[, c("Depression", "Study Satisfaction", "Job Satisfaction", "Sleep Duration")]
```
Used summary instead of describe()

```{r}
# Defining skewness and kurtosis
Skewness <- function(x) {
  n <- length(x)
  m3 <- sum((x - mean(x))^3)/n
  s3 <- sd(x)^3
}

kurtosis <- function(x) {
  n <- length(x)
  m4 <- sum((x - mean(x))^4)/n
  s4 <- sd(x)^4
  m4/s4
}


summary(StudentDep)
```
```{r}
# Calculating mean, Standard Deviation, Skewness, and Kurtosis
Vals <- sapply(StudentDep, is.numeric)
Data <- StudentDep[, Vals]

# Removing NA
Data <- Data[, colSums(is.na(Data)) < nrow(Data)]

SumStats <- data.frame(
  variable = colnames(Data),
  Mean = sapply(Data, mean),
  SD = sapply(Data, sd),
  Skewness = sapply(Data, Skewness),
  Kurtosis = sapply(Data, kurtosis)
)
print(SumStats)
```
Finding Covariance, Correlation, Scatter Plot Matrices. A issue faced would be that R is taking binary number that uses for example; Yes = 1, No = 0, into the correlation/covariance which is causing NA to be put into the dataset.

```{r}
# Covariance Matrix
Covar <- cov(Data)
print("Covariance Matrix:")
```
```{r}
Covar
```
```{r}
# Correlation Matrix
Corr <- cor(Data)
print("Correlation Matrix:")
Corr
```

```{r}
# Scatter Plot
pairs(Data, pch = 19, main = "Scatter Plot")
```
```{r}
# Correlation Plot
corrplot(cor(Data), method = "square")
```
The “???” are NA correlation values, this is because some of the rows can have the same values. We would need to remove, so we can get a better understanding on the correlation.

```{r}
pairs(Data, pch = 19, main = "Scatter matrix for outliers")
```

```{r}
# Mahalanobis distance
Vals <- sapply(Data, is.numeric)
Data <- Data[, Vals]

# Remove
Data <- na.omit(Data)

# Remove columns that have zero variance
Data <- Data[, apply(Data, 2, var) != 0]

center <- colMeans(Data)
Covar <- cov(Data)

mahalDist <- mahalanobis(Data, center, Covar)
Data$mahalanobis <- mahalDist


# Identify Outliers
CutOff <- qchisq(0.975, df = ncol(Data) - 1)
Outliers <- which(mahalDist > CutOff)
Outliers
```

When using the Mahalanobis distance and the use of scattterplots, we wanted to observe the correlation matrix and the outliers. We saw that certain variables may be weaker or stronger which depends between the variables. Also, can see that some extreme data points misshaped the strengths as well the binary variables caused the dataset to not look as appealing.